The natural history of scavenging in vertebrates

### 2016-03-18 natural history of scavenging.tex
added refs

@@

Avian flight originates in the Jurassic Period, conincident with the fossils of \textit{Archaeopteryx lithographica} so many of these benefits would have been realised from that point on for carnivorous birds. However, vertebrate flight is much older than this where pterosaurs predate bird origins by a considerable margin in the Triassic Period. Scavenging in this diverse group has been hypothesied many times before. Certain clades of these animals could reach enormous sizes (e.g. Azhdarchids with wingspans of 11 metres) and look to have engaged in soaring flight \citep{witton2010size}.  Although \cite{witton2008reappraisal} argued that neck inflexibility and straight, rather than hooked jaw morphology points against their existing as \textit{obligate} scavengers, Azhdarchid terrestrial proficency indicates they would have been comfortable foraging on the ground. Indeed, extant Marabou Storks have a comparable morphology and are noted facultative scavengers so it is reasonable to believe that certain pterosaurs behaved similarly.

Large body size confers substantial dominance and starvation resistance benefits \citep{ruxton2004obligate}. Thus, we would expect scavengers to have this trait selected for even in the case of weight-constrained fliers. Cinereous Vultures (\textit{Aegypius monachus}) and condors (\textit{Vultur gryphus}, \textit{Gymnogyps californianus}) all have body masses that can exceed 10 kg and represent some of the heaviest bird species capable of flight \citep{ferguson2001raptors,donazar2002effects}. And as we have noted the Azhdarchid pterosaurs were far bigger again, with estimated body masses of around 80 kg.

The only other vertebrate group capable of powered flight are the bats where scavenging has not been recorded to our knowledge. Their visual acuity is famously poor and echolocation does not lend itself to discovering immobile carrion. Their small size and poor terrestrial ability would also count against them at a carcass. The bat fossil record is notoriously poor owing to their fragile skeletons so we are unable to determine if extinct species were more suited to this lifestyle.

\section*{Terrestrial Scavengers}
A simplificaiton of terrestrial, vertebrate scavengers is one of them existing in a two-dimensional plane while foraging for carrion directly. They can detect carcasses at a range that is defined by the radius of their sensory organs, usually the visual and olfactory senses. As a consequence, they have a much more restricted view of the landscape than do aerial foragers. No contemporary terrestrial vertebrate exists as an obligate scavenger but most if not all are facultative scavengers to some extent. \cite{ruxton2004obligate} offer a reason for this in that the traits that allow for vultures to exist as scavengers undermined their ability to hunt but that the same forces have not prevented mammals from doing so. The same authors calculated that "a 1 tonne mammal or reptile, in an ecosystem yielding carrion at densities similar to the current Serengeti, could have met its energy requirements if it could detect carrion over a distance of the order of 400–500 m."\citep{ruxton2004obligate}.

@@

The order Carnivora sees its origins in the Eocene where it split into the Caniforma and Feliforma. And many of the aforesaid adapations for scavenging are found in these other major terrestrial mammalian carnviores. Though the specific mix of features realised in hyenas suggest this is the model organism for terrestrial scavenging among mammals in the past. For example, the bone-crushing dogs that evolved during the Oligocene (subfamily Borophaginae) have been compared to hyenas in terms of their feeding ecology \citep{van2003chapter,martin2016pursuit}. Certainly, it is profitable to compare extant terrestrial species to their prehistoric forebears given the dominance of mammalian carnviores since the Eocene.

Interestingly such comparisons have given insight into the feeding ecology of early hominins who, for instance, had the ability to craft tools for breaking open bones \citep{hone2010feeding,ARCM:ARCM12084}. The question of where our ancestors placed on the hunter-scavenger axis during the Plio-Pleistocene has been a matter of debate for years. A recent study investigating potential scavenging opportunities for hominins in Kenya found that, even when discounting bone material, there is a substantial amount of scavengeable meat left on predated remains; sufficient to sustain the requirements of an adult male \textit{Homo erectus} \citep{pobiner2015new}. In some historical hominin-inhabited areas there were a greater number of felids than hyenids. This is significant because hyenas will leave far less flesh on a carcass than a felid such as a sabretooth enabling contemperaneous hominins to benefit \citep{pobiner2015new}. The intelligence, resultant tool-use and cooperative nature of hominins meant they could likely adapt to take on more or less carrion depending on their environment.

@@

As we discussed for the case of Cenozoic carnivores, osteophagy could be extremely beneficial to a scavenger. 
In Mesozoic systems some extremely large theropod dinosaurs had a morphology which suggests an ability to process bone e.g. the robust skull and dentition of \textit{T rex}. There is direct evidence that \textit{T.rex} did this in the form of distinctive wear marks on its tooth apices \citep{farlow1994wear,schubert2005wear} and the presence of bone fragments in its coprolites \citep{chin1998king}. The animal also had an enormous bite force, with one estimate putting it at 57000 Newtons \citep{bates2012estimating}. This is noted as being powerful enough to break open skeletal material \citep{rayfield2001cranial}. Osteophagy may have been even more viable during this era because the body mass distribution of herbviores tended to be skewed towards larger sizes \citep{10.1371/journal.pone.0051925}. When we couple this with the fact that skeletal mass scales greater than linearly with body mass \citep{prange1979scaling} there would have been a lot of bone material to consume in the environment provided an animal had the biology to process it  \citep{chure1997one}.

@@

\section*{Aquatic Scavengers} 
An aquatic environment presents challenges for direct observational studies and so, similar to the approaches involving extinct species, much work has approached the question of scavenging propensity from an energetics perspective. Although as with the aerial and terrestrial enviornments we have evidence of facultative scavenging among aquatic species. For example, the remains of a mosasaur and a terrestrial hadrosaur were discovered with embedded teeth from a Cretaceous shark \textit{Squalicorax} \citep{schwimmer1997scavenging}. As well as a likely instance of scavenging between a 4-million-year-old white shark (\textit{Carcharodon}) and mysticete whale from Peru \citep{ehret2009caught}.
The existence of an obligate scavenger in a marine setting is uncertain \citep{britton1994marine,smith2003ecology,ruxton2004energetic,ruxton2005searching}. Depending on the species a carcass in this environment either floats or descends to the sea floor. In the latter low-light environment, detection distances are far lower (< 100 m) than they would be in the air. As such, animals detect resources through chemo- and mechanoreception more so than through vision \citep{ruxton2004energetic}. However, water is a medium that is conducive to low-cost movement \citep{tucker1975energetic} and so may be able to support an obligate scavenging fish \citep{ruxton2004energetic,ruxton2005searching}. \cite{benbow2015introduction} do note that "some benthic scavengers (e.g., hagfish: family Myxinidae) rely on necrophagy for a large portion of their diet and may indeed be obligate scavengers".

@@

% Slight changes in temperature could have profound impacts on energy flow within ecosystems as carrion decomposition
% rates roughly double for every 10°C increase in ambient temperature (Vass et al. 1992; Parmenter and
% MacMahon 2009). Consequently, the availability of carrion to vertebrate scavengers could be reduced by
%as much as 20%–40% over the next century based on projections from current climate change models
% (Beasley et al. 2012b).

%"Changes in global climate are also expected to substantially alter the availability of carrion resources in marine ecosystems as surface production of organic material could decline by 50\% or more (Smith et al. 2008). Such a drastic reduction in productivity undoubtedly will impact a multitude of ecosystem processes, including the availability and distribution of carrion within marine ecosystems." \citep{benbow2015introduction}

%"For example, vertebrate scavengers appear to be disadvantaged when humidity and temperature favor microbial and invertebrate reproduction (between 21C and 38C)" \citep{benbow2015introduction}

%"In a study by DeVault et al. (2004b), it was found that at temperatures above 20C vertebrates were able to detect and consume only 19\% of small-mammal carcasses, whereas at temperatures below 18C, vertebrates consumed 49\% of carcasses" \citep{benbow2015introduction}

%"For example, the available literature suggests that carcass decomposition occurs more rapidly in warm, moist habitats when compared with more xeric habitats" \citep{benbow2015introduction}

%Of particular note is the increase in decomposition rate of carrion with temperature which could result in a seasonal reduction of carrion availability given predicted warming trends \citep{moleon2014inter}.

%"Apart from being an important water-conservation strategy, nocturnal behavior may have evolved in the Hyaenidae as a means of reducing competition with the other dominant scavengers in African ecosystems, the vultures, which are exclusivly diurnal (Houston 1979)" Carnivore Behavior, Ecology, and Evolution By John L. Gittleman

% "Thus, on an annual basis, the FMR (and food requirements) of an endotherm are likely to be around 30 times greater than those for an equivalent-sized ectotherm"

%"Maximal sustainable speeds of mammals are six- to sevenfold those of lizards of equal size,"

%Scavenging is a widespread behaviour among vertebrates where most if not all carnivores act as facultative scavengers.
%TG: or something more along the lines as:
%Most if not all carnivorous vertebrates are facultative scavenging behaviour to some extant.
%It is recognised that scavengers have an important role in keeping energy flows at a higher trophic level in food webs than decomposers because they consume relatively more carrion \citep{devault2003scavenging}. 
%Scavengers also provide useful ecosystem services by acting as barriers to the spread of disease by quickly consuming rotting carcasses which have often died from illness \citep{ogada2012dropping}.
%(Since we are intrested in scavanging in th paleo record ecosystem services might not be that relavent, although modulators of disease is still relavent) 
%Despite this, scavengers are a seriously understudied group \citep{sekercioglu2006increasing,selva2007nested,wilson2011scavenging}.  
%\cite{devault2003scavenging} propose that this is due to both human disgust at carrion itself and the difficulty in determining if an ingested prey item was killed or scavenged. 
%The latter point means that studying the natural history of this behaviour is particularly problematic.
%Indeed, even data on the proportion of carrion in the diet of extant species are sorely lacking \citep{benbow2015introduction}.

%"Indeed, in some ecosystems, vertebrates have been documented to assimilate as much as 90\% of the available carrion"\citep{benbow2015introduction}.

@@

%Our review is divided up into three sections, namely the land, air and sea. 
%These are then subdivided into three geological eras the Cenozoic, Mesozoic and Palaeozoic. 
%Each of these environments has a distinct phyiscal character that affects how a species forages for food which has obvious relevance for an animal searching for carrion.  
%However, there is some commonality to a scavenger's environment and the problems in finding food that one would encounter. 
%Notably, the resource environment of a scavenger is a patchily distributed one, because it is difficult to predict both when and where a carcass is produced. 
%As a result of this, any animal existing as a scavenger must maximise its detection capabilities and minimise its locomotory costs \citep{ruxton2004obligate}.
%Exploring which groups are likely to have moved towards these traits and thus existed as sccavengers over palaentological time forms the basis of this work.
%remember the journal is an ecological one at heart so I would always keep in mind about making it useful for ecologists over paelo people. In particular from the website "ECOGRAPHY publishes papers focused on broad spatial and temporal patterns, particularly studies of population and community ecology, macroecology, biogeography, and ecological conservation. Studies in ecological genetics and historical ecology are welcomed in the context of explaining contemporary ecological patterns".

%"Interkingdom Competition among Vertebrates, Invertebrates, and Microbes"\citep{benbow2015introduction}
%"For example, vertebrate scavengers appear to be disadvantaged when humidity and temperature favor microbial and invertebrate reproduction" \citep{benbow2015introduction}

@@

% "Changes in global climate are also expected to substantially alter the availability of carrion resources in marine ecosystems as surface production of organic material could decline by 50% or more (Smith et al. 2008). Such a drastic reduction in productivity undoubtedly will impact a multitude of ecosystem processes, including the availability and distribution of carrion within marine ecosystems." \citep{benbow2015introduction}

%\subsection*{Cenozoic}
%Among terrestrial African carnivores, hyenas, jackals, lions and leopards all take sizable proportions of carrion in their diet.
%In the case of the spotted hyena (\textit{Crocuta crocuta}) it can be as high as 99\% \citep{benbow2015introduction}. 
%Yet, no contemporary terrestrial vertebrate exists as an obligate scavenger. 
%The selective pressures that push mammals and reptiles towards scavenging do not seem to undermine their ability to hunt, perhaps explaining the absence of obligate scavengers in these groups \citep{ruxton2004obligate}.
%\\There is a long running debate on the tendency of human ancestors to scavenge. 
%Some recent studies have found "that passive scavenging from abandoned larger felid kills could have been a high-yield, though potentially dangerous, foraging strategy for early hominins even without considering within-bone nutrients" \citep{pobiner2015new}.

%Osteophagy is known across a range of terrestrial carnivores.
%Some fat-rich mammalian bones have an energy density (6.7 kJ/g) comparable with that of muscle tissue, making skeletal remains an enticing resource \citep{brown1989study}. 
% Hyenas have a bite force capable of breaking open bones and 
% In light of this, the skeletal remains of carrion may act as trove of food to carnivores who can access it.

@@

% A review by \cite{devault2002scavenging} found occurrences of scavenging spread across five families of snakes and states that this behaviour is "far more common than currently acknowledged."

%"\cite{sazima1990necrofagia} predicted that  several components  of  snake  be-havior contribute to scavenging propensity. They  postulated  that  snakes  that  rely  on chemosensory information to acquire prey might forage  for  carrion  more  frequently than  those  whose  modes  of prey acquisition are driven by visual cues. Additionally, \cite{sazima1990necrofagia} suggested that habitat and diet preferences might be indicators of scavenging   propensity   be-cause  they  influence  the  likelihood of a particular species encountering carrion naturally. Based on these criteria, they pre-dicted that aquatic or semi-aquatic pisciv-orous snakes  would  scavenge  more  fre-quently than other species. Moreover, \cite{sazima1990necrofagia}  suggested that water currents that induce  aggrega-tions of carrion heighten the probability of carrion detection by these snakes (see also \cite{savitzky1992laboratory}).  Also,  chemical  gradients may  be  more  uniform  and  give  more  de-pendable  directional information in water \citep{sazima1990necrofagia}." \citep{devault2002scavenging}

%"Investigations of ophidian metabolic re-quirements unveil additional advantages to carrion  utilization.  Snakes  exhibit  exceed-ingly  low  maintenance  metabolisms,  and most  can  survive  on  a  few  scant  feedings per year. It  is, therefore, possible for snakes to rely large-ly  on  infrequent,  less  energy-rich  meals. Carrion, which is by nature ephemeral and unpredictable, may represent such a food source  to  snakes.  Scavenging  might  allow snakes to meet their low metabolic needs more  easily  and  without  the  costs  associ-ated with subduing prey." \citep{devault2002scavenging}

%\textit{Allosaurus} tooth marks on a hadrosaur in the Late Jurassic. 
% Late Triassic scavenging on a prosauropod by basal carnivorous archosaurs \citep{hungerbuhler1998taphonomy}.

%\subsection*{Palaeozoic}
% The absence of flying vertebrates in the Palaeozoic may have permitted terrestrial forms to take in a higher proportion of carrion in their diet.

%\subsection*{Cenozoic}
%A likely instance of scavenging between a 4-million-year-old white shark (\textit{Carcharodon}) and mysticete whale from Peru \citep{ehret2009caught}.
%Bite marks on early Holocene Tursiops truncatus fossils from the North Sea indicate scavenging by rays (Chondrichthyes, Rajidae) \citep{van2009bite}. 
%Possible scavenging on a juvenile fur seal from the Late Neogene \citep{boessenecker2011mammalian}.

%\subsection*{Mesozoic}

%\subsection*{Palaeozoic}
%Evolution of sharks, known scavengers.

%\section*{Results}

%\section*{Discussion}

>>>>

%TG: what do you mean by their position? Phylogenetic? Position on the "scavenging scale"?

@@

Avian flight originates in the Late Jurassic (163.5-145 Million years ago - Mya), conincident with the fossils of \textit{Archaeopteryx lithographica} so many of these benefits would have been realised from that point on for carnivorous birds. However, vertebrate flight is much older than this where pterosaurs predate bird origins by a considerable margin in the Late Triassic (235-201.3 Mya). Scavenging in this diverse group has been hypothesied many times before. Certain clades of these animals could reach enormous sizes (e.g. Azhdarchids with wingspans of 11 metres \citep{witton2010size})%CITE
 and look to have engaged in soaring flight \citep{witton2010size}. Although \cite{witton2008reappraisal} argued that neck inflexibility and straight, rather than hooked jaw morphology points against their existing as \textit{obligate} scavengers, Azhdarchid terrestrial proficency indicates they would have been comfortable foraging on the ground. Indeed, extant Marabou Storks (\textit{Leptoptilos crumenifer}) %TG: Check if that's the one
have a comparable morphology and are noted facultative scavengers so it is reasonable to believe that certain pterosaurs behaved similarly.

Large body size confers substantial dominance and starvation resistance benefits \citep{ruxton2004obligate}. Thus, we would expect scavengers to have this trait selected for even in the case of weight-constrained fliers. Cinereous Vultures (\textit{Aegypius monachus}) and condors (\textit{Vultur gryphus}, \textit{Gymnogyps californianus}) all have body masses that can exceed 10 kg and represent some of the heaviest bird species capable of flight \citep{ferguson2001raptors,donazar2002effects}. And as we have noted the Azhdarchid pterosaurs were far bigger again, with estimated body masses of over 200 kg \citep{witton2010size}.  %CITE

The only other vertebrate group capable of powered flight are the bats where scavenging has not been recorded to our knowledge. Their visual acuity is famously poor and echolocation would not lend itself to discovering immobile carrion. Their small size and poor terrestrial ability would also count against them at a carcass \citep{riskin2006terrestrial}. The bat fossil record is notoriously poor owing to their fragile skeletons so we are unable to determine if extinct species were more suited to this lifestyle \citep{eiting2009global}.
%TG: Two points here: one speculation for the non-scavenging behaviour could be because of diet. Big bats (which are better suited for scavenging, following previous paragraph) are essentially frugivores and therefore probably lack of adaptations for digesting the meat. Also carnivores bats are mainly the microbats that are insectivores. Maybe they scavenge on dead insects but here we're talking about scavenging on real meat right?
%TG: Second point: bats fossil record is of course not so good but it's actually way better than birds! Most of the data we have is on teeth of course but it's probably enough for having a rough body mass and diet estimate! And that's much better than birds where you're stuck with just some fragments of long bones.

\section*{Terrestrial Scavengers}
A simplification of terrestrial, vertebrate scavengers is one of them existing in a two-dimensional plane while foraging for carrion directly. They can detect carcasses at a range that is defined by the radius of their sensory organs, usually the visual and olfactory senses. As a consequence, they have a much more restricted view of the landscape than do aerial foragers. No contemporary terrestrial vertebrate exists as an obligate scavenger but most if not all are facultative scavengers to some extent. \cite{ruxton2004obligate} offer a reason for this in that the traits that allow for vultures to exist as scavengers undermined their ability to hunt but that the same forces have not prevented mammals from doing so. The same authors calculated that "a 1 tonne mammal or reptile, in an ecosystem yielding carrion at densities similar to the current Serengeti, could have met its energy requirements if it could detect carrion over a distance of the order of 400–500 m."\citep{ruxton2004obligate}.

@@

The order Carnivora sees its origins in the Middle Eocene (56-33.9 Mya) where it split into the Caniforma and Feliforma. And many of the aforesaid adapations for scavenging are found in these other major terrestrial mammalian carnviores. Though the specific mix of features realised in hyenas suggest this is the model organism for terrestrial scavenging among mammals in the past. For example, the bone-crushing dogs that evolved during the Oligocene (subfamily Borophaginae) have been compared to hyenas in terms of their feeding ecology \citep{van2003chapter,martin2016pursuit}. Certainly, it is profitable to compare extant terrestrial species to their prehistoric forebears given the dominance of mammalian carnviores since the Eocene.

Interestingly such comparisons have given insight into the feeding ecology of early hominins who, for instance, had the ability to craft tools for breaking open bones \citep{hone2010feeding,ARCM:ARCM12084}. The question of where our ancestors placed on the hunter-scavenger axis during the Plio-Pleistocene has been a matter of debate for years. A recent study investigating potential scavenging opportunities for hominins in Kenya found that, even when discounting bone material, there is a substantial amount of scavengeable meat left on predated remains; sufficient to sustain the requirements of an adult male \textit{Homo erectus} \citep{pobiner2015new}. In some historical hominin-inhabited areas there were a greater number of felids than hyenids. This is significant because hyenas are likely to have left far less flesh on a carcass than a felid such as a sabretooth enabling contemperaneous hominins to benefit \citep{pobiner2015new}. 
%TG: or "could"? This is hypothetical behaviour stuff! Maybe past hyenas where exclusive sushi eaters!
%TG: Not sure about the intelligence thing: the sentence as the same meaning without and will be less anthropocentric sounding.
The intelligence, resultant tool-use and cooperative nature of hominins meant they could likely adapt to take on more or less carrion depending on their environment.

@@

As we discussed for the case of Cenozoic carnivores, osteophagy could be extremely beneficial to a scavenger. 
In Mesozoic systems some extremely large theropod dinosaurs had a morphology which suggests an ability to process bone e.g. the robust skull and dentition of \textit{T. rex}. There is direct evidence that \textit{T. rex} did this in the form of distinctive wear marks on its tooth apices \citep{farlow1994wear,schubert2005wear} and the presence of bone fragments in its coprolites \citep{chin1998king}. The animal also had an enormous bite force, with one estimate putting it at 57000 Newtons \citep{bates2012estimating}. This is noted as being powerful enough to break open skeletal material \citep{rayfield2001cranial}. Osteophagy may have been even more viable during this era because the body mass distribution of herbviores tended to be skewed towards larger sizes \citep{10.1371/journal.pone.0051925}. When we couple this with the fact that skeletal mass scales greater than linearly with body mass \citep{prange1979scaling} there would have been a lot of bone material to consume in the environment provided an animal had the biology to process it \citep{chure1997one}.

@@

\section*{Aquatic Scavengers}

An aquatic environment presents challenges for direct observational studies and so, similar to the approaches involving extinct species, much work has approached the question of scavenging propensity from an energetics perspective. Although as with the aerial and terrestrial enviornments we have evidence of facultative scavenging among aquatic species. For example, the remains of a mosasaur and a terrestrial hadrosaur were discovered with embedded teeth from a Cretaceous shark \textit{Squalicorax} \citep{schwimmer1997scavenging}. As well as a likely instance of scavenging between a 4-million-year-old white shark (\textit{Carcharodon}) and mysticete whale from Peru \citep{ehret2009caught}.
The existence of an obligate scavenger in a marine setting is uncertain \citep{britton1994marine,smith2003ecology,ruxton2004energetic,ruxton2005searching}. Depending on the species a carcass in this environment either floats or descends to the sea floor. In the latter low-light environment, detection distances are far lower (< 100 m) than they would be in the air. As such, animals detect resources through chemo- and mechanoreception more so than through vision \citep{ruxton2004energetic}. However, water is a medium that is conducive to low-cost movement \citep{tucker1975energetic} and so may be able to support an obligate scavenging fish \citep{ruxton2004energetic,ruxton2005searching}. \cite{benbow2015introduction} do note that "some benthic scavengers (e.g., hagfish: family Myxinidae) rely on necrophagy for a large portion of their diet and may indeed be obligate scavengers".

@@

%Scavenging is a widespread behaviour among vertebrates where most if not all carnivores act as facultative scavengers.
%TG: or something more along the lines as:
%Most if not all carnivorous vertebrates are facultative scavenging behaviour to some extant.
%It is recognised that scavengers have an important role in keeping energy flows at a higher trophic level in food webs than decomposers because they consume relatively more carrion \citep{devault2003scavenging}. 
%Scavengers also provide useful ecosystem services by acting as barriers to the spread of disease by quickly consuming rotting carcasses which have often died from illness \citep{ogada2012dropping}.
%(Since we are intrested in scavanging in th paleo record ecosystem services might not be that relavent, although modulators of disease is still relavent)

@@

%remember the journal is an ecological one at heart so I would always keep in mind about making it useful for ecologists over paelo people. In particular from the website "ECOGRAPHY publishes papers focused on broad spatial and temporal patterns, particularly studies of population and community ecology, macroecology, biogeography, and ecological conservation. Studies in ecological genetics and historical ecology are welcomed in the context of explaining contemporary ecological patterns".

@@

@@

%\textit{Allosaurus} tooth marks on a hadrosaur in the Late Jurassic. 
% Late Triassic scavenging on a prosauropod by basal carnivorous archosaurs \citep{hungerbuhler1998taphonomy}.
%A likely instance of scavenging between a 4-million-year-old white shark (\textit{Carcharodon}) and mysticete whale from Peru \citep{ehret2009caught}.
%Bite marks on early Holocene Tursiops truncatus fossils from the North Sea indicate scavenging by rays (Chondrichthyes, Rajidae) \citep{van2009bite}. 
%Possible scavenging on a juvenile fur seal from the Late Neogene \citep{boessenecker2011mammalian}.
